\section{Introduction}
\label{Sec:Introduction}
Reachability analysis is a foundational technique providing a rigorous and systematic means of analyzing software with respect to specified safety properties.
%Model checking is a fundamental part of software verification. 
While it is undecidable in general, recent advances in Satisfiability Modulo Theories (SMT) and Constrained Horn Clause (CHC) solvers 
have enabled the development of efficient symbolic analysis techniques. 
However, when the program of interest has a complicated loop structure, 
the search space for possible property violations explodes quickly.
Thus, the reachability analysis remains challenging even for advanced modern techniques. 

%
State-of-the-art techniques such as 
Counter-Example Guided Abstraction Refinement (CEGAR)~\cite{DBLP:conf/cav/ClarkeGJLV00,DBLP:conf/fmcad/HojjatR18,DBLP:conf/fmcad/BeyerKW10},
IC3/Spacer~\cite{DBLP:journals/fmsd/KomuravelliGC16,DBLP:journals/fac/BradleyM08},  
Accelerated Bounded Model Checking (ABMC)~\cite{DBLP:journals/corr/abs-2401-09973}, 
and Transition Power Abstraction (TPA)~\cite{DBLP:conf/tacas/BlichaFHS22,DBLP:conf/fm/BritikovBSF24} 
 often struggle with nested loops and complex control flows. 
Although capable of some acceleration, approaches based on IC3/Spacer and CEGAR usually perform reachability analysis one step at a time, which is particularly taxing in the presence of deep loops.
%They can be inefficient for reachability analysis that involves traversing deep loops.
A recent \tpa technique instead makes larger steps and handles deep loops more efficiently. 
However, it applies only to the analysis of a single loop. 
To analyze nested loops, \tpa relies on a preprocessing to merge them into a single loop, losing important 
structural information. 
ABMC underapproximates loop behavior and is more suited for bug finding than for proving safety. 
Specifically, ABMC attempts to generate blocking clauses to prove safety, but it can be challenging for programs with nested loops and complex execution flow.

%

We introduce a new approach of Cycle Summarization-based Reachability Analysis (CSRA) to target programs with nested loops and complex structures. 
Our algorithm operates within the framework of CHC solving, in which the program is modeled by logical implications over uninterpreted predicates. 
CHCs serve as an intermediate software representation, which enables reasoning about programs independently of their programming language. 
This paradigm is employed in a wide range of model checkers, such as 
\textsc{SeaHorn}~\cite{DBLP:conf/cav/GurfinkelKKN15}, \textsc{JayHorn}~\cite{DBLP:conf/cav/KahsaiRSS16}, 
\textsc{RustHorn}~\cite{DBLP:conf/esop/0002T020}, and \textsc{Korn}~\cite{DBLP:conf/tacas/Ernst23}.

%

Our approach decomposes reachability analysis into the exploration of different cycle-free paths 
through the CHC system. 
The algorithm traverses each path independently, analyzing cycles and constructing cycle 
unrollings when needed. 
The core procedure of our approach is the computation and refinement of cycle summaries. 
\solveCHC uses \tpa engine \emph{modularly}, to analyze cycles, produce cycle summaries 
and unrollings. 
%The use of \tpa also enables efficient deep reachability checks. 
The constructed cycle summaries serve as abstractions that describe any possible cycle 
unrolling and as components for the construction of other cycle summaries. 
% This allows inner cycle summaries to be used in the construction of the overapproximation 
% of any possible cycle iteration. 
The use of cycle summaries enables efficient reasoning about potentially infinite sets of 
paths in the system and facilitates safety verification. 
Additionally, once learned, these summaries can be reused in the analysis of other paths 
through the CHC system. 
%\solveCHC enables efficient handling of programs with nested loops, as demonstrated by our experiments.

%

A prototype of our technique has been implemented in the \textsc{Golem}~\cite{DBLP:conf/cav/BlichaBS23} CHC solver and 
evaluated on benchmarks from CHC-Comp-2024~\cite{DBLP:journals/corr/abs-2404-14923}, 
aeval-bench-marks~\cite{DBLP:conf/cav/FedyukovichPMG19}, and a set of instances synthesized from 
real C++ code from repositories of \textsc{Z3}~\cite{DBLP:conf/birthday/BjornerGMR15}, \textsc{SeaHorn}~\cite{DBLP:conf/cav/GurfinkelKKN15}, 
\textsc{OpenSMT}~\cite{DBLP:conf/sat/HyvarinenMAS16} and \textsc{Golem}. 
Our results show that the approach is competitive with other state-of-the-art reachability 
analysis techniques. 
Specifically, \solveCHC solved a significant number of instances uniquely.
On average, \solveCHC is 2.5$\times$ faster than 
\textsc{Eldarica} and 1.9$\times$ faster than \textsc{Spacer}. 

The main contributions of this paper are: (1) a novel CHC-based reachability analysis approach 
for nested-loop programs, (2) a new TPA-based compositional approach for the generation and refinement of cycle summaries
that uses inner cycle summaries to produce outer cycle summaries, 
(3) a technique for decomposing CHC reachability analysis into the exploration of cycle-free paths through the CHC system, 
(4) formalization and proof of soundness for our method, and 
(5) an extensive empirical evaluation of CSRA in comparison with other state-of-the-art approaches for reachability analysis.


% is a novel CHC-based reachability analysis algorithm 
% for nested-loop programs. 
% Our method introduces a compositional approach for constructing and refining cycle 
% summaries within the CHC system, enabling efficient safety verification. 
% Modular usage of TPA allows the algorithm to find deep counterexamples and prove the 
% safety of the system using the constructed summaries. 
% Additionally, decomposing reachability analysis into exploration of separate simple paths 
% simplifies the verification process and creates potential for the 
% parallelization. \konst{Make it into items}


% \konst{Write a paragraph here}
% \begin{itemize}
%     \item A novel CHC-solving algorithm for the verification of nested-loop programs.
%     \item A compositional cycle summarization and refinement procedure.
% \end{itemize}


\subsection*{Motivating example}

Figure~\ref{fig:ex} gives a C program with a nested loop structure. 
The outer loop increments \texttt{x1} by the value of \texttt{x2} and assigns 0 to \texttt{x2}, 
while the inner loop increments \texttt{x2} by 2 until \texttt{x2} $\geq$ 32. 
When execution of the loops is finished, \texttt{x1} equals 512, so the program is safe.

% 
% Rename looping clause to inductive clause
% 
\begin{subfigures}
\begin{figure}[t!]
\begin{lstlisting}[language=C++,label={lst:multi},escapechar=|]
int x1 = 0, x2 = 0;
while (x1 < 512) { |\label{line:Cloop1}|
 while(x2 < 32) { x2 = x2 + 2; } |\label{line:Cloop2}|
 x1 = x1 + x2; x2 = 0;
}
assert(x1 == 512);
\end{lstlisting}
	\caption{A C program with nested loops.}
	\label{fig:ex}
\end{figure}
%

\begin{figure}[t!]
    \centering
    {
        \scriptsize
        \begin{align*}
            &x_1=0 \land x_2 = 0 &&\implies \pred_1(x_1,x_2)&&(\clause_1)\\
            &\pred_1(x_1,x_2) \land x_1 < 512 \land x_1' = x_1 \land x_2'=x_2 &&\implies \pred_2(x_1', x_2')&&(\clause_2)\\
            &\pred_2(x_1,x_2) \land x_2 < 32 \land x_1' = x_1 \land x_2'=x_2+2 &&\implies \pred_2(x_1', x_2')&&(\clause_3)\\
            &\pred_2(x_1,x_2) \land x_2 \geq 32 \land x_1' = x_1+x_2 \land x_2'=0 &&\implies \pred_1(x_1', x_2')&&(\clause_4)\\
            &\pred_1(x_1,x_2) \land x_1 \geq 512 \land x_1 \neq 512 &&\implies \bot&&(\clause_5)
        \end{align*}
    }

    \caption{CHC representation of the program in Figure~\ref{fig:ex}.}
    \label{fig:chcs}
\end{figure}


\end{subfigures}
 
When applied to this program, state-of-the-art reachability analysis techniques behave as follows.
In the worst-case scenario, CEGAR and Spacer/IC3 would need to traverse all 16 outer and 256 inner loop iterations one by one to 
synthesize \emph{inductive invariants} for both inner and outer loops. 
In contrast, TPA would proceed using \emph{transition invariant} generation, but in order to apply it, one would need to merge both loops into a single one, losing structural information of the program and thus degrading efficiency. 
For this specific example, ABMC would underapproximate inner and outer loop behavior, successfully deriving 
blocking clauses that prove the safety of the program. 
However, as shown by our evaluation, it often struggles, especially if the control flow is more complex or the program contains multiple nested loops.

%


In contrast, \solveCHC leverages the program's structural information to efficiently prove safety 
and handle deep loops. 
First, \solveCHC constructs a cycle summary that overapproximates the 
behavior of the inner loop. 
Then, using this summary as an abstraction of any possible inner loop traversal, \solveCHC employs
\tpa to analyze the outer loop and prove the safety of the system. 
The use of the inner cycle summary eliminates the need to produce exact traces through the loop, 
making our approach significantly more efficient compared to classical \tpa.