\section{Termination Analysis via Interpolation-based Transition Invariant Generation}

Our main procedure for proving termination leverages safety verification to generate a well-founded transition invariant via interpolation.
At each iteration, the algorithm constructs a trace from the initial states to a sink state, used for the termination proof.
The algorithm tracks states covered by the disjunctively well-founded transition invariant, focusing on those not yet covered.
The interpolation-based approach enables property-directed transition invariant generation, focusing on conditions relevant to termination.

\subsection{Interpolation-based Generation of Well-Founded Transition Invariants}

The core of our approach is the automated, interpolation-based generation of well-founded transition invariants, and the refinement of 
the state space with respect to the states covered by the transition invariant.
Goal is to have an heuristic, which would allow to overapproximate the transition relations, and use the constructed overapproximation 
as a foundation for the transition invariant (as the core property of transition invariant is that it is an overapproximation of reflexive 
transitive closure).

\konst{Move to introduction later.} State of the art approaches that attempt to construct transition invariants, are centered
around the usage of template-bbased synthesis[citations needed]. 
This significantly limits the capabilities of transition-invariant generation, as it is restricted only to the templates.
Additionally, this template-based generation is somewhat random, not guided by the conditions necessary for the termination.
Proposed interpolation-based transition invariant generation allows to direct those two issues.


The procedure consists of four steps: (1) computing $T^n$, the set of states guaranteed to terminate within $n$ steps; 
(2) constructing an over-approximation of the terminating trace via interpolation; 
(3) filtering the resulting candidates for well-foundedness; and 
(4) verifying that the candidate transition invariant covers all reachable states.

First, it constructs a formula $T^n$ encoding all states that are guaranteed to reach sink states 
(terminate) within $n$ transitions:
$$\forall x^{(0)}. T^{n}(x^{(0)}) \implies \forall x^{(1)}, \dots, x^{(n)}. Tr^{n}(x^{(0)}, \dots, x^{(n)}) \implies Sink(x^{(n)})$$

$T^n$ is constructed with quantifier elimination, eliminating all variables from the formula except for $x^{(0)}$:

$$T^{n}(x^{(0)}) = \lnot QE(\exists x^{(1)}, \dots, x^{(n)} | Tr^{n}(x^{(0)}, \dots, x^{(n)}) \land \lnot Sink(x^{(n)}))$$

Construction of such states enables the construction of an overapproximation of the transition relation, guided by the sink states and focusing on 
the variables relevant to termination. 
We construct an over-approximation of the trace via Craig interpolation. The following formula is unsatisfiable by construction:

$$T^{n}(x^{(0)}) \land Tr^{n}(x^{(0)}, \dots, x^{(n)}) \land \lnot Sink(x^{(n)})$$

This unsatisfiability allows us to compute the interpolant $Cands(x, x') = Itp(Tr^{n}(x^{(0)}, \dots, x^{(n)}), T^{n}(x^{(0)}) \land \lnot Sink(x^{(n)}))$, 
which over-approximates the trace.
The key intuition is that the usage of the sink states for interpolant construction allows the interpolant to capture the reason \textbf{why} 
the particular trace is guaranteed to lead to the termination.
Next, the algorithm uses this interpolant as a foundation for the construction of the transition invariant.
For transition invariant to be a proof of termination, it needs to be disjunctively well-founded.
Our technique, first converts $Cands(x,x')$ to a Disjunctive Normal Form (DNF): $Cands(x,x') = \bigvee_i d_i$.
The number of disjuncts is bounded by a constant (a tunable parameter depending on the available size of RAM), since DNF 
conversion can cause an exponential blowup in formula size.

The procedure then filters the candidate disjuncts.
For each disjunct $d(x,x') \in Cands(x,x')$ algorithm checks if $d(x,x')$ corresponds to a well-founded relationvia automated linear ranking function synthesis~\cite{DBLP:conf/vmcai/PodelskiR04}:

$$TrInv(x,x') = \bigvee_i \{d_i \in Cands| \text{for $d_i \exists$ linear ranking function $f(x,x')$} \}$$

If $TrInv$ is a transition invariant over all reachable states, then TS terminates, since every disjunct corresponds to well-founded relation.
To verify this, procedure computes the formula $C(x)$ encoding the states over which $TrInv$ is a valid transition invariant. 
Such states holds a following property:

$$\forall x, x',x''. C(x) \land (TrInv(x,x') \lor Id(x,x')) \land Tr(x',x'') \implies TrInv(x,x'')$$

Similarly to $T^{n}$, this formula can be constructed using the quantifier elimination, negating the states for 
which $TrInv$ is not a transition invariant:

$$C(x) = \lnot QE(\exists x',x'' | (TrInv(x,x') \lor Id(x,x')) \land Tr(x',x'') \land \lnot TrInv(x,x''))$$

The algorithm then performs a safety check of a problem $P = \tuple{Init, Tr, \lnot C, x}$, to determine whether $\lnot C(x)$ 
is reachable from $Init$.
If $\lnot C(x)$ is not reachable, then $TrInv(x,x')$ is a disjunctively well-founded transition invariant over all reachable
states, witnessing the termination of transition system.
Otherwise, if $\lnot C(x)$ is reachable, the algorithm commences a refinement procedure, which is described in more detail below.

\begin{lemma}
    Let $T = \tuple{Init, Tr, x}$ be a transition system and let 
    $C(x)$ be a formula such that $\lnot C(x)$ is unreachable in 
    $T$. If $TrInv(x,x')$ is a union of well-founded 
    relations and satisfies:
    $$\forall x,x',x''.\ C(x) \land (TrInv(x,x') \lor Id(x,x')) 
      \land Tr(x',x'') \implies TrInv(x,x'')$$
    then $T$ is terminating.
\end{lemma}
\begin{proof}
    We show that $TrInv$ is a disjunctively well-founded transition 
    invariant over the reachable states of $T$, from which 
    termination follows by~\cite{DBLP:conf/lics/PodelskiR04}.

    \textbf{Well-foundedness.} By construction, each disjunct 
    $d_i$ of $TrInv = \bigvee_i d_i$ admits a linear ranking 
    function~\cite{DBLP:conf/vmcai/PodelskiR04}, so each $d_i$ is 
    a logical encoding of well-founded relation. 
    Hence $TrInv$ is disjunctively well-founded.

    \textbf{Transition invariant condition.} We show that 
    $\forall x,x'.\ R(x) \land Tr^*(x,x') \implies TrInv(x,x')$,
    where $R(x)$ denotes the reachable states of $T$, by induction 
    on the number of steps $n$ in $Tr^*(x,x')$.

   \emph{Base case} ($n = 0$): $Tr^0(x,x') \equiv Id(x,x')$. 
    Since $\lnot C(x)$ is unreachable, $R(x) \implies C(x)$. 
    Instantiating the hypothesis with $Id(x,x')$:
    $$C(x) \land Id(x,x') \land Tr(x',x'') \implies TrInv(x,x'')$$
    Simplifying the formula since $Id(x,x')$ is an identity relation gives 
    $C(x) \land Tr(x,x') \implies TrInv(x,x')$, and since 
    $R(x) \implies C(x)$, the base case holds.

   \emph{Inductive step}: Assume $R(x) \land Tr^n(x,x') \implies 
    TrInv(x,x')$ for some $n \geq 0$. 
    Let $Tr^{n+1}(x,x'')$ hold, so there exists $x', x''$ with $R(x) \land Tr^n(x,x') \land Tr(x',x'')$. By 
    the inductive hypothesis, $TrInv(x,x')$ holds. 
    Since $R(x) \land Tr^n(x,x') \implies R(x) \land TrInv(x,x')$, then
    $R(x) \land Tr^n(x,x') \land Tr(x',x'') \implies R(x) \land TrInv(x,x') \land Tr(x',x'')$.
    Now, due to the fact that $R(x) \implies C(x)$, the following formula is implied
    $C(x) \land TrInv(x,x') \land Tr(x',x'')$.
    And by claim in the lemma, $C(x) \land TrInv(x,x') \land Tr(x',x'') \implies TrInv(x,x'')$.

    Therefore:
    $$R(x) \land Tr^{n+1}(x,x') \implies TrInv(x,x')$$

    Thus $TrInv$ is a disjunctively well-founded transition 
    invariant over all reachable states, and $T$ is 
    terminating~\cite{DBLP:conf/lics/PodelskiR04}.
\end{proof}

\subsection{Algorithm Description}


The algorithm description is given in Algorithm~\ref{Alg:SNA_Term}. 
Below, we explain the steps of the algorithm in detail.
Then, provide an algorithm execution example.
Afterwards, we prove the correctness of the introduced approach.

Our technique, similarly to the Algorithm~\ref{Alg:SNA} takes as input a Safety Problem $P=\tuple{Init, Tr, Sink, x}$, where $Init$ and 
$Tr$ encode the initial states and transition relation, $Sink$ is the formula encoding sink states, and $x$ is 
a set of variables over which the Transition System is defined.
As an output, the algorithm returns NONTERM if TS is nonterminating, TERM if it is terminating, and UNKNOWN if the procedure cannot determin the final result.



\begin{algorithm}[t!]
	\SetInd{0.5em}{0.5em}
    \DontPrintSemicolon
    \small
    \newcommand{\Continue}{\textbf{continue}}
    \caption{Termination analysis via interpolation-based Transition Invariant generation.}
    \label{Alg:SNA_Term}
    \SetKwInOut{Input}{Input}
    \SetKwInOut{Output}{Output}
    \Input{Safety problem $P = \tuple{Init, Tr, Sink, x}$}
    \Output{TERM | NONTERM}
    $TInv \gets \bot$ \\
    $Init' \gets Init$ \\
    \While{$\top$} {
        \label{SNA_Term:Safety} \Switch{$safetyCheck(\tuple{Init', Tr, Sink, x})$}{
            \Case{$\tuple{UNSAFE, Tr^{n}(x^{(0)}, \dots, x^{(n)})}$} {
                $\tuple{Init', Tr, Sink, x} \gets \blockTerm(Tr, Tr^{n})$
                \label{SNA_Term:T} $T^n(x) \gets \lnot QE(\exists x^{(1)}, \dots, x^{(n)} | Tr^{n}(x, x^{(1)}, \dots, x^{(n)}) \land \lnot Sink(x^{(n)}))$  \\
                $Cs(x,x') \gets DNFIZE(Itp(Tr^{n}(x, x^{(1)}, \dots, x'), T^n(x) \land \lnot Sink(x')))$ \label{SNA_Term:Itp}\\
                $TInv \gets TInv \lor \bigvee_i d_i$ \label{SNA_Term:TINV} \tcp*[r]{$\bigvee_i d_i$ are well-founded disjuncts in $Cs$} 
                $C(x) \gets \lnot QE(\exists x',x''|$ $Tr(x,x') \land (Id(x',x'') \lor TInv(x',x'')) \land \lnot TInv(x,x''))$\label{SNA_Term:NC} \\
                \label{SNA_Term:CheckNc} \Switch{$safetyCheck(\tuple{Init, Tr, \lnot C, x})$}{ 
                    \Case{$\tuple{UNSAFE, Tr^{n}(x^{(0)}, \dots, x^{(m)})}$} {
                        % $\tuple{Init, Tr, Sink, x} \gets BlockTerm(\tuple{Init', Tr, Sink, x}, Tr^{n})$ \\ 
                        $M \models Init(x^{(0)}) \land Tr^{n}(x^{(0)}, \dots, x^{(m)}) \land Sink(x^{(m)})$ \label{SNA_Term:UNSAFE}  \\
                        $Init'(x) \gets $ \textit{variable assignment extracted from} $M$ for $x^{(m)}$\\
                    }      
                    \Case{$\tuple{SAFE, Inv(x)}$} {
                        \eIf {$Init' \neq Init$} 
                        { 
                            $Init' \gets Init$
                        }{
                            \Return $TERM$\label{SNA_Term:NEWTERM}
                        }
                    } 
                }
            }
            \Case{$\tuple{SAFE, Inv(x)}$} {
            %    \label{SNA_TERM:SAFE}
                \lIf{$\forall x. Inv(x) \rightarrow \exists x'. Tr(x,x') \land Inv(x')$} {
                    \Return $NONTERM$ \label{SNA_TERM:NONTERM}
                }
                \Return UNKNOWN
            }
        }
    }
\end{algorithm}
% Return BlockTerm to algorithm maybe!!!!
% Use relations when I need to define W-F
% Don't use relations in the algo description

The procedure iterates until it either finds a recurrent set - proving nontermination, or a disjunctively well-founded Transition Invariant, 
that allows to conclude the termination.
During the whole algorithm, the procedure constructs transition invariant, progressively extending the transition invariant by adding well-founded disjuncts.
Additionally, it updates the initial states, proving termination for particular reachable states in the Transition system.
Every iteration, similarly to the Algorithm~\ref{Alg:SNA}, begins with the safety check at line~\ref{SNA_Term:Safety}.
The algorithm handles two possible cases:

\paragraph{UNSAFE.} It means there exists a terminating trace in the Transition System. 
At the lines~\ref{SNA_Term:T}-\ref{SNA_Term:CheckNc} algorithm executes the interpolation-based generation of transition invariant 
candidate, following the approach described above - generating $T^n$, constructing an overappproximation of transition trace and
transforming it into union of well-founded relations.

% $$\forall x. T^n(x) \implies \forall x^{(1)}, \dots, x^{(n)}. trace(x,x^{(1)}, \dots, x^{(n)}) \implies Sink(x^{(n)})$$

% This formula can be constructed by applying the quantifier elimination (QE) on the second part of the formula:
% $\lnot QE(\exists x^{(1)}, \dots, x^{(n)} | trace(x, x^{(1)}, \dots, x^{(n)}) \land \lnot Sink(x^{(n)}))$. 
% Afterwards, algorithm proceeds to use $T^n$ to construct an overapproximation of the terminating trace, using the interpolation over:

% $$T^n(x) \land trace(x, x^{(1)}, \dots, x') \land \lnot Sink(x')$$

% This formula is unsatisfiable by construction, as $T^n$ cannot reach states outside of $Sink(x')$ after n transitions, and this formula is used for 
% the interpolation at the line~\ref{SNA_Term:CheckNc}.
% After the overappproximation of trace is constructed, it is transformed into disjunctive normal form. 
% Any possible technique for dnfization can be used.
% At the line~\ref{SNA_Term:TINV} our algorithm filters disjuncts, adding well-founded disjuncts from the interpolant to the transition invariant candidate.
% The filtering is done based on the synthesis of the linear ranking function~\cite{DBLP:conf/lics/PodelskiR04}.
% If it is possible to synthesise a linear real function for specific disjunct, it is added to the $TInv$, otherwise it is dropped.
% After the filtering, algorithm constructs the states that are not covered by the current transition invariant candidate.
% The core property for transition invariant is\footnote{In the formula below, $Id(x',x'') : x' = x''$ is an identity formula preserving the variable values}:
% $$Tr(x,x') \land (Id(x',x'') \lor TInv(x',x'')) \implies TInv(x,x'')$$ 

% Based on this it is possible to construct formula encoding all states that are not covered by the current transition invariant candidate:
% $$\forall x. Nc(x) \implies \lnot \forall x',x'' Tr(x,x') \land (Id(x',x'') \lor TInv(x',x'')) \implies TInv(x,x'')$$

% At the line~\ref{SNA_Term:NC} algorithm constructs the formula $Nc(x)$ using quantifier elimination.
% Afterwards, our technique checks if $Nc(x)$ is reachable.
If it is possible to reach state in $\lnot C(x)$ from the initial states, then our approach tries to prove termination of this state, not covered by 
transition invariant, by updating $Init'(x)$ to this state.
Otherwise, if $C(c)$ is not reachable from the original Initial states $Init(x)$ - algorithms concludes that TS terminates.
If $C(x)$ is not reachable from the updated Initial states $Init'(x) \neq Init(x)$, then algorithm resets to the original initial states, and attemtps to prove termination with an updated transition invariant. 


\paragraph{SAFE.} In case $Sink$ states are not reachable, the algorithm terminates.
In this case the invariant is a recurrent set by default, as the transition formula is not changed.
Therefore, the transition system is nonterminating.

\begin{example}
    Consider a terminating transition system $T$ in which:
    \begin{itemize}
        \item $Init: \top$
        \item $Tr(x,x'): (v' = v - 1 \lor v' = v - 2) \land v > 0$
        \item $x: \{v\}$
    \end{itemize}

    Initially, before the algorithm execution, preprocessing constructs the sink states: 
    $$Sink(x) = QE(\exists x'|\lnot Tr(x,x')) = v \leq 0.$$
    For any state with $v \leq 0$ it is impossible to do a transition in the TS.
    Additonally, algorithm would update initial states: 
    $$Init'(x) = Init(x) \land \lnot Sink(x) = v > 0.$$
    Since every state that satisfies sink states will terminate by default.

    %

    The algorithm is called with $P=\tuple{Init', Tr, Sink, x}$.
    First, it executes the safety check, detecting the trace that leads to the sink states.
    For example, it detects a trace of length 2:  $ Init(x) \land Tr(x,x') \land Tr(x',x'') \land Sink(x)$.
    Then, it first constructs $T^n(x)$, formula for all of the states that are guaranteed to terminate in one or less transitions:
    $$T^n(x) = v \leq 2$$
    Then, it constructs an interpolant:
    $$Itp(Tr(x,x^{(1)}) \land Tr(x^{(1)},x'), T^n(x) \land Sink(x')) = v' \leq v - 1 \land v' > 0.$$

    This interpolant is a conjunction, therefore the dnfization returns the same formula in the line~\ref{SNA_Term:Itp}.
    Using the linear ranking function synthesis method, it is proven that $v' \leq v - 1 \land v > 0$ is well-founded, and the 
    transition invariant candidate $TInv(x,x') = v' \leq v - 1 \land v > 0$.
    At the line~\ref{SNA_Term:NC}, algorithm constructs states that are not covered by the invariant:
    $$Nc(x) = v \leq 1$$
    The safety check then detects that $Nc(x)$ is not reachable from the initial states, therefore transition invariant candidate
    $TInv$ covers all the reachable states.
    Since $TInv$ is well-founded, algorithm concludes that $T$ terminates.
\end{example}


\subsection{Correctness}

This subsection proves the soundness of the introduced approach. We first prove that when algorithm returns TERM, the transition system is nonterminating,
and then we prove the correctnessof the nontermination.


\begin{lemma}
    If Algorithm~\ref{Alg:SNA_Term} returns $TERM$, then the transition system $\tuple{Init, Tr, x}$ is terminating.
\end{lemma}
\begin{proof}
    Algorithm~\ref{Alg:SNA_Term} returns $TERM$ only at the line~\ref{SNA_Term:NEWTERM}.
    If algorithm returns $TERM$ then the safety check at line~\ref{SNA_Term:CheckNc} returns SAFE.
    It implies that $Nc(x)$ is unreachable from $Init(x)$ under $Tr'$. 
    Additionally, by construction, $Nc(x)$ captures all states for which $TInv$ is not a transition invariant.
    It means that for states satisfying $\lnot Nc(x)$ $TInv$ is a transition invariant under $Tr'$:
    $$\forall x,x',x''. \lnot Nc(x) \land (TInv(x,x') \lor Id(x,x')) \land Tr'(x',x'') \land \lnot Nc(x'') \rightarrow TInv(x,x'')$$
    By construction, $TInv$ is also disjunctively well-founded (line~\ref{SNA_Term:TINV}).

    Since (1) $Nc(x)$ is not reachable from $Init(x)$ based on the check at line~\ref{SNA_Term:CheckNc}, and (2) $TInv$ is a disjunctively well-founded transition invariant over all remaining states
     - the transition system $\tuple{Init, Tr, x}$ is terminating~\cite{DBLP:conf/lics/PodelskiR04}.
\end{proof}


\begin{lemma}
    If Algorithm~\ref{Alg:SNA_Term} returns $NONTERM$, then the transition system $\tuple{Init, Tr, x}$ is nonterminating.
\end{lemma}
\begin{proof}
    The proof of this lemma follows directly from the lemma~\ref{Lem:NONTERM}. 
    To prove it we prove the existence of the recurrent set, for which two conditions should hold:
    \begin{itemize}
        \item $\forall x. N(x) \rightarrow \exists x'. Tr(x,x') \land N(x')$
        \item $\exists x. N(x) \land Init(x)$
    \end{itemize}
    \emph{First condition:} For $Inv(x)$ the following holds: $\forall x. Inv(x) \rightarrow \exists x'. Tr(x,x') \land Inv(x')$.
    This is the case as two properties below hold:
    \begin{itemize}
        \item $\lnot Sink(x) \implies \exists x'. Tr(x,x')$ (by construction of $Sink(x)$)
        \item $Inv(x) \implies \lnot Sink(x)$ (As a result of safety check)
    \end{itemize}

    \emph{Second condition.} For any safety problem $\tuple{Init', Tr, Sink}$, $Init'$ is reachable from the initial states $Init$ by construction.
    The states that are reached on the trace from $Init$ to $Init'$ can be added to $Inv(x)$ to construct a full recurrent set.

    Therefore, there exists a recurrent set within the transition system, and it is nonterminating.
\end{proof}


% \begin{algorithm}[t!]
% 	\SetInd{0.5em}{0.5em}
%     \DontPrintSemicolon
%     \small
%     \newcommand{\Continue}{\textbf{continue}}
%     \caption{Analyze(P)}
%     \label{Alg:SNA_Term}
%     \SetKwInOut{Input}{Input}
%     \SetKwInOut{Output}{Output}
%     \Input{Safety problem $P = \tuple{Init, Tr, Sink, x}$}
%     \Output{$\tuple{NONTERM, R(x)}$ | $\tuple{TERM, TInv(x,x')}$}
%     \While{$\top$} {
%         $TInv \gets \bot$ \\
%         \Switch{$safetyCheck(\tuple{Init, Tr, Sink, x})$}{
%             \Case{$\tuple{UNSAFE, trace(x^{(0)}, \dots, x^{(n)})}$} {
%                 $T^n(x), \forall x. T^n(x) \rightarrow \forall x^{(1)}, \dots, x^{(n)}. trace(x, x^{(1)}, \dots, x^{(n)}) \rightarrow Sink(x^{(n)})$  \\
%                 $Cs(x,x') \gets DNFIZE(Itp(trace(x, x^{(1)}, \dots, x'), T^n(x) \land \lnot Sink(x')))$ \\
%                 $TInv \gets TInv \lor \bigvee_i d_i$\tcp*[r]{$\bigvee_i d_i$ are w-f disjuncts in $Cs$} 
%                 $Nc(x) \gets \exists x',x''. Tr(x,x') \land (Id(x',x'') \lor TInv(x',x'')) \land \lnot TInv(x,x'')$ \\
%                 \Switch{$safetyCheck(\tuple{Init, Tr, Nc, x})$}{ 
%                     \label{SNA_Term:CheckNc}
%                     \Case{$\tuple{SAFE, Inv(x)}$} {
%                          \Return $\tuple{TERM, TInv(x,x')}$\label{SNA_Term:NEWTERM}
%                     }
%                     \Case{$\tuple{UNSAFE, trace(x^{(0)}, \dots, x^{(m)})}$} {
%                         $M \models Init(x^{(0)}) \land trace(x^{(0)}, \dots, x^{(m)}) \land Sink(x^{(m)})$ \label{SNA_Term:UNSAFE}  \\
%                         $s_m(x) \gets $ \textit{variable assigment extracted from } $M$ for $x^{(m)}$\\
%                         \Switch{$Analyze(\tuple{s_m(x), Tr, Sink, x})$} {
%                             \Case{$\tuple{NONTERM, R(x)}$} {
%                                 \Return $\tuple{NONTERM, R(x)}$ \label{SNA_Term:NEWNONTERM}
%                             }
%                             \Case{$\tuple{TERM, TInv'(x,x')}$} {
%                                 $TInv(x,x') \gets TInv(x,x') \lor TInv'(x,x')$ \\
%                             }
%                         }
%                     }       
                
%                 }
%             }
%             \Case{$\tuple{SAFE, Inv(x)}$} {
%                \label{SNA_TERM:SAFE}
%                 \lIf{$\forall x. Inv(x) \rightarrow \exists x'. Tr(x,x') \land Inv(x')$} {
%                     \Return $\tuple{NONTERM, Inv(x)}$ \label{SNA_TERM:OLDNONTERM}
%                 }
%                 \Return UNKNOWN
%             }
%         }
%     }
% \end{algorithm}

% Algorithm uses the terminating trace to generate states $T^n$, that necessarily lead to termination in up to $n$ steps of transition relation.
% Then, using interpolation, the algorithm construct an over-approximation of the $n \leq i \leq 1$ steps of transition relation.
% This interpolant is then transformed into a Disjunctive Normal Form (DNF), and every disjunct is separately checked for well-foundness, using linear Ranking function synthesis~\cite{DBLP:conf/lics/PodelskiR04}.
% The disjunction of all well-founded disjuncts is then added to the candidate transition invariant formula $TInv$.

% \begin{lemma}
%     If Algorithm~\ref{Alg:SNA_Term} returns $\tuple{TERM, TInv(x,x')}$, then the transition system $\tuple{Init, Tr, x}$ is terminating.
% \end{lemma}
% \begin{proof}
%     Algorithm~\ref{Alg:SNA_Term} returns $\tuple{TERM, TInv(x,x')}$ only at the line~\ref{SNA_Term:NEWTERM} or line~\ref{SNA_Term:OLDTERM}.
%     In the first case, the algorithm returns $\tuple{TERM, TInv(x,x')}$.
%     By construction, $TInv(x,x')$ is a disjunction of well-founded relations, each synthesised via linear ranking functions, so $TInv$ is disjunctively well-founded.

%     It remains to show that $TInv$ is a transition invariant over all reachable states. 
%     The safety check at line~\ref{SNA_Term:CheckNc} returning SAFE implies that $Nc(x)$ is unreachable from $Init(x)$ under $Tr'$. 
%     Additionally, by construction, $Nc(x)$ captures all states for which $TInv$ is not a transition invariant.
%     It means that for states satisfying $\lnot Nc(x)$ $TInv$ is a transition invariant under $Tr'$:
%     $$\forall x,x',x''. \lnot Nc(x) \land (TInv(x,x') \lor Id(x,x')) \land Tr'(x',x'') \land \lnot Nc(x'') \rightarrow TInv(x,x'')$$

%     In the formula $Tr'(x,x') = Tr(x,x') \land \lnot \bigvee_i s_i(x')$, where $\bigvee_i s_i(x')$ are the states that have 
%     been blocked during the analysis and are guaranteed to terminate.
%     $TInv$ is and therefore is a disjunctively well-founded transition invariant over the states described by $\lnot Nc(x) \land \lnot \bigvee_i s_i(x')$.

%     Since (1) $Nc(x)$ is not reachable from $Init(x)$ based on the check at line~\ref{SNA_Term:CheckNc}, (2) states in $\bigvee_i s_i(x')$ are guaranteed to terminate, and (3) $TInv$ is a transition invariant over all remaining states
%      - the transition system $\tuple{Init, Tr, x}$ is terminating~\cite{DBLP:conf/lics/PodelskiR04}.

%     In the second case, the proof follows directly from Lemma~\ref{Lem:TrivialTermination}, as it is based on the same check of $Init(x) \land Tr(x,x')$.
% \end{proof}


% \begin{lemma}
%     If Algorithm~\ref{Alg:SNA_Term} returns $\tuple{NONTERM, R(x)}$, then the transition system $\tuple{Init, Tr, x}$ is nonterminating.
% \end{lemma}
% \begin{proof}
%     Algorithm~\ref{Alg:SNA_Term} returns $\tuple{NONTERM, R(x)}$ at the line~\ref{SNA_Term:NEWNONTERM} or line~\ref{SNA_Term:OLDNONTERM}.
%     At line~\ref{SNA_Term:OLDNONTERM}, the algorithm returns $\tuple{NONTERM, Inv(x)}$, where $Inv(x)$ is a recurrent set, which implies that the transition system is nonterminating (following from the Lemma~\ref{Lem:NontermPreservation}).
    
%    Another case is when the algorithm returns the result of a recursive call $Analyze(\tuple{s_m(x), Tr, Sink, x})$. 
%    By the inductive hypothesis applied to this call, $R(x)$ is a recurrent set for $\tuple{s_m, Tr, x}$, satisfying:
%     \begin{enumerate}[label=\Roman*.]
%         \item $\exists x.\ s_m(x) \land R(x)$, and
%         \item $\forall x.\ R(x) \rightarrow \exists x'. Tr'(x,x') \land R(x')$.
%     \end{enumerate}
%     Since $s_m(x)$ is reachable from $Init(x)$ (line~\ref{SNA_Term:CheckNc}), $R(x)$ is also reachable from $Init(x)$, and therefore $R(x)$ is a recurrent set for the original transition system, which implies that the transition system is nonterminating.
% \end{proof}


% \begin{algorithm}[t!]
% 	\SetInd{0.5em}{0.5em}
%     \DontPrintSemicolon
%     \small
%     \newcommand{\Continue}{\textbf{continue}}
%     \caption{Analyze(P)}
%     \label{Alg:SNA_Term}
%     \SetKwInOut{Input}{Input}
%     \SetKwInOut{Output}{Output}
%     \Input{Safety problem $P = \tuple{Init, Tr, Sink, x}$}
%     \Output{$\tuple{NONTERM, R(x)}$ | $\tuple{TERM, TInv(x,x')}$}
%     \While{$\top$} {
%         $TInv \gets \bot$ \\
%         \Switch{$safetyCheck(\tuple{Init, Tr, Sink, x})$}{
%             \Case{$\tuple{UNSAFE, trace(x^{(0)}, \dots, x^{(n)})}$} {
%                 -- SAME AS IN ALGORITHM 2 -- \\
%                 $T^n(x), \forall x. T^n(x) \rightarrow \forall x^{(1)}, \dots, x^{(n)}. trace(x, x^{(1)}, \dots, x^{(n)}) \rightarrow Sink(x^{(n)})$  \\
%                 $Cs(x,x') \gets DNFIZE(Itp(trace(x, x^{(1)}, \dots, x'), T^n(x) \land \lnot Sink(x')))$ \\
%                 $TInv(x,x') \gets TInv(x,x') \lor \bigvee_i d_i$\tcp*[r]{$\bigvee_i d_i$ are w-f disjuncts $\in Cs$} 
%                 $Nc(x) \gets \exists x',x''. Tr(x,x') \land (Id(x',x'') \lor TInv(x',x'')) \land \lnot TInv(x,x'')$ \\
%                 \Switch{$safetyCheck(\tuple{Init, Tr, Nc, x})$}{ 
%                     \label{SNA_Term:CheckNc}
%                     \Case{$\tuple{SAFE, Inv(x)}$} {
%                          \Return $\tuple{TERM, TInv(x,x')}$\label{SNA_Term:NEWTERM}
%                     }
%                     \Case{$\tuple{UNSAFE, trace(x^{(0)}, \dots, x^{(m)})}$} {
%                         $M \models Init(x^{(0)}) \land trace(x^{(0)}, \dots, x^{(m)}) \land Sink(x^{(m)})$ \label{SNA_Term:UNSAFE}  \\
%                         $s_m(x) \gets $ \textit{variable assigment extracted from } $M$ for $x^{(m)}$\\
%                         \Switch{$Analyze(\tuple{s_m(x), Tr, Sink, x})$} {
%                             \Case{$\tuple{NONTERM, R(x)}$} {
%                                 \Return $\tuple{NONTERM, R(x)}$ \label{SNA_Term:NEWNONTERM}
%                             }
%                             \Case{$\tuple{TERM, TInv'(x,x')}$} {
%                                 $TInv(x,x') \gets TInv(x,x') \lor TInv'(x,x')$ \\
%                             }
%                         }
%                     }       
                
%                 }
%             }
%             \Case{$\tuple{SAFE, Inv(x)}$} {
%                  \label{SNA_Term:SAFE}
%                 \lIf {UNSAT ? $Init(x) \land Tr(x,x')$} {
%                     \Return $\tuple{TERM, TInv(x,x')}$\label{SNA_Term:OLDTERM}
%                 }
%                 \If{$\forall x. Inv(x) \rightarrow \exists x'. Tr(x,x') \land Inv(x')$} {
%                     \Return $\tuple{NONTERM, Inv(x)}$ \label{SNA_Term:OLDNONTERM}
%                 } 
%                 $Sink(x) \gets \forall x. Sink'(x) \rightarrow \lnot \exists x'. Tr(x,x')$
%             }
%         }
%     }
    
% \end{algorithm}

\section{Safety-based Nontermination Analysis Extensions}
\label{Sec:Main}

In this section, we present our Transition System-based extensions for safety-based termination analysis, which allow to prove termination of transition systems and to improve the nontermination analysis. 


\subsection{Extension of the Algorithm for Termination of Transition Systems}

Algorithm~\ref{Alg:SNA} can be extended to prove termination of transition systems via a single additional SMT check.
Specifically, when the algorithm reaches the SAFE case at line~\ref{SNA:SAFE}, the following check is performed:
$$Init(x) \land Tr(x,x')$$
This check is intended to determine whether it is possible to take a transition after all traces to sink states have been blocked.
Throughout the execution of the algorithm, the transition relation is progressively refined to exclude states that deterministically lead to sink states.
If this formula is unsatisfiable, then the transition system is terminating (since Init states lead to sink states).


\subsection{Safety Problem Refinement}

The algorithm can further refine the safety problem $P$ in the case where $Inv$ does not constitute a recurrent set.
Particularly, it is possible to extract states within the invariant, from which it is impossible to take a transition (which makes $Inv$ not a recurrent set).

$$\forall x. Sink'(x) \land \lnot \exists x'. Tr'(x,x')$$

In this formula, $Sink'(x)$ is a new formula which describes sink states within the inductive invariant.
It can be computed  by applying quantifier elimination to $\lnot \exists x'. Tr'(x,x')$.
Afterwards, the produced formula $Sink'(x)$ replaces $Sink(x)$ in the safety problem $P$, enabling further analysis.
After those two updates the algorithm would look as follows:
%% USE If/Else instead of Switch/Case
%% Define the function for the Trace Block
%% Present a scenario when we go to TrInvCand is a Transition Invariant and returns YES, then it is terminating - no refinement
%% Do a loop - no matter how many times I loop - still terminating
%% Try to in

\begin{algorithm}[t!]
	\SetInd{0.5em}{0.5em}
    \DontPrintSemicolon
    \small
    \newcommand{\Continue}{\textbf{continue}}
    \caption{Nontermination analysis using Safety.}
    \label{Alg:SNA_Refined}
    \SetKwInOut{Input}{Input}
    \SetKwInOut{Output}{Output}
    \Input{Safety problem $P = \tuple{Init, Tr, Sink, x}$}
    \Output{NONTERM | TERM}
    \While{$\top$}{
        \Switch{$safetyCheck(\tuple{Init, Tr, Sink, x})$}{
            \Case{$\tuple{UNSAFE, trace(x^{(0)}, \dots, x^{(n)})}$} {
                -- SAME AS IN ALGORITHM 1 -- 
            }
            \Case{$\tuple{SAFE, Inv(x)}$} {
                \label{SNA_Refined:SAFE}
                \lIf {UNSAT ? $Init(x) \land Tr(x,x')$} {
                    \Return TERM \label{SNA_Refined:TERM}
                }
                \lIf{$\forall x. Inv(x) \rightarrow \exists x'. Tr(x,x') \land Inv(x')$} {
                    \Return NONTERM \label{SNA_Refined:RECURRENT}
                } 
                $Sink(x) \gets \forall x. Sink'(x) \rightarrow \lnot \exists x'. Tr(x,x')$
            }
        }
    }
    
\end{algorithm}


This algorithm differs from the original one in two lines.
Particularly, at line~\ref{SNA_Refined:TERM} the algorithm checks if one can identify states within the invariant from which no transition is possible — precisely those states that prevent $Inv$ from being a recurrent set.
It allows to check the termination in some cases.
And at line~\ref{SNA_Refined:SAFE} the algorithm refines the sink states, by extracting states from the invariant, from which it is impossible to take a transition.
Which allows to proceed with termination analysis and allows to solve more complex problems.


\begin{lemma}\label{Lem:TrivialTermination}
    If the formula $Init(x) \land Tr(x,x')$ is unsatisfiable at line~\ref{SNA_Refined:TERM} of Algorithm~\ref{Alg:SNA_Refined}, then the transition system $\tuple{Init, Tr, x}$ is terminating.
\end{lemma}
\begin{proof}
    The algorithm maintains the following invariants throughout its execution:
    \begin{align}
        Tr'(x,x')   &= Tr(x,x') \land \textstyle\bigwedge_i \lnot s_i(x') \label{eq:TrRefined}\\
        Init'(x)    &= Init(x)  \land \textstyle\bigwedge_j \lnot s_j(x) \label{eq:InitRefined}
    \end{align}
    where each blocked state $s_i(x')$ (resp. $s_j(x)$) is one that deterministically leads to a sink state, and is therefore guaranteed to terminate if reached.

    Suppose $Init'(x) \land Tr'(x,x')$ is unsatisfiable.  
    Expanding the definitions gives:
    $$
      \bigl(Init(x) \land \textstyle\bigwedge_j \lnot s_j(x)\bigr) 
      \;\land\; 
      \bigl(Tr(x,x') \land \textstyle\bigwedge_i \lnot s_i(x')\bigr)
      \;\equiv\; \bot
    $$
    This means that every initial state from which a transition is still possible has been blocked, i.e., every reachable initial state either cannot take a transition, or can only reach a state in $\{s_i\}$ or originate from a state in $\{s_j\}$, all of which are guaranteed to terminate. 
    Since no infinite trace can be initiated from any such state, the transition system is terminating.
\end{proof}

\begin{lemma}\label{Lem:NontermPreservation}
    If the refined transition system $\tuple{Init', Tr', x}$ is nonterminating, then the original transition system $\tuple{Init, Tr, x}$ is also nonterminating.
\end{lemma}
\begin{proof}
    Suppose $\tuple{Init', Tr', x}$ is nonterminating. 
    Then there exists a recurrent set $Inv'(x)$ such that:
    \begin{enumerate}[label=\Roman*.]
        \item $\exists x.\ Init'(x) \land Inv'(x)$, and
        \item $\forall x.\ Inv'(x) \rightarrow \exists x'.\ 
              Tr'(x,x') \land Inv'(x')$.
    \end{enumerate}
    We show that $Inv'(x)$ is also a recurrent set for the original system $\tuple{Init, Tr, x}$.

    \textbf{Reachability from $Init$.} Since $Init'(x) = Init(x) \land \bigwedge_i \lnot s_i(x)$, we have $Init'(x) \rightarrow Init(x)$. 
    Combined with condition~(I), there exists a state $x_0$ such that $Init(x_0) \land Inv'(x_0)$ holds, so $Inv'$ is reachable from $Init$.

    \textbf{Closure under $Tr$.} Since $Tr'(x,x') = Tr(x,x') \land \bigwedge_i \lnot s_i(x')$, we have $Tr'(x,x') \rightarrow Tr(x,x')$. 
    Therefore, for any $x$ with $Inv'(x)$, condition~(II) guarantees the existence of $x'$ with $Tr'(x,x') \land Inv'(x')$, and hence $Tr(x,x') \land Inv'(x')$.

    Thus $Inv'(x)$ satisfies both conditions of a recurrent set for $\tuple{Init, Tr, x}$, which is therefore nonterminating.
\end{proof}

