\section{Related Work}
\label{Sec:Related}
Termination and nontermination analysis are active research areas. 
Below we discuss the techniques related to our approach.


\textbf{Nontermination.} 
Nontermination analysis is typically centered on the construction 
of recurrent sets — sets of states from which execution cannot 
escape~\cite{DBLP:conf/cade/FrohnG22,DBLP:conf/tacas/BrockschmidtCIK16,
DBLP:conf/tacas/ChenCFNO14,DBLP:conf/foveoos/BrockschmidtSOG11,
DBLP:conf/cav/LarrazNORR14,FedyukovichZG18}.
These techniques range from template-based synthesis to 
safety-check-based recurrent set construction, the latter of which 
our approach directly extends~\cite{DBLP:conf/tacas/ChenCFNO14}.
An alternative approach detects nontermination by identifying 
lasso-shaped executions — finite paths leading to a cycle — via 
bounded model checking~\cite{DBLP:journals/entcs/BiereAS02}.
Leike and Heizmann~\cite{DBLP:conf/tacas/LeikeH18} propose 
geometric nontermination witnesses: finite representations of 
infinite executions via sums of geometric series.
On the other hand, our technique is integrated with termination 
analysis, so that intermediate results from each analysis 
guide the other.

%  within a unified framework
% Nontermination analysis is typically centered around the construction of recurrent sets — sets of states from which execution cannot 
% escape~\cite{DBLP:conf/sas/Ben-AmramDG19,DBLP:conf/cade/FrohnG22,DBLP:conf/tacas/BrockschmidtCIK16,DBLP:conf/tacas/ChenCFNO14,DBLP:conf/foveoos/BrockschmidtSOG11,DBLP:conf/cav/LarrazNORR14}.
% These techniques can vary from the template-based synthesis to the safety-check based recurrent set construction.
% Another traditional approach is safety-based nontermination analysis~\cite{DBLP:journals/entcs/BiereAS02},
% that detects nontermination if the same state is reachable within TS more then once.
% Another interesting approach is the construction of geometric nontermination argument~\cite{DBLP:conf/tacas/LeikeH18} -
% a finite representation of an infinite execution that via a sum of geometric series.

\textbf{Termination.} 
The most widely studied approach to termination analysis is the 
synthesis of ranking functions~\cite{DBLP:conf/vmcai/PodelskiR04,
DBLP:conf/cav/KuraUH20,DBLP:journals/jacm/Ben-AmramG14,
DBLP:journals/jar/GieslABEFFHOPSS17,DBLP:conf/esop/SaritaSGSV26,
DBLP:conf/tacas/UrbanGK16,FedyukovichZG18,RileyF25}.
State-of-the-art techniques focus on the construction of 
multiphase or lexicographic ranking 
functions~\cite{DBLP:journals/jacm/Ben-AmramG14,FedyukovichZG18}, which 
significantly broaden the class of programs for which termination 
can be proved. However, the complexity of a transition system can 
make ranking function synthesis infeasible.
An alternative is fixpoint computation~\cite{DBLP:journals/pacmpl/UnnoTGK23}, 
which reduces termination and nontermination to the validity of a first-order fixpoint formula.
Another family of techniques proves termination via 
disjunctively well-founded transition 
invariants~\cite{DBLP:conf/lics/PodelskiR04,DBLP:conf/sas/CookPR05,
DBLP:conf/cav/KroeningSTW10,DBLP:conf/tacas/TsitovichSWK11}.
These approaches over-approximate the behavior of the transition 
system, reducing the termination proof to a disjunctive 
well-foundedness check on the invariant. 
The central challenge is the generation of the 
transition invariant itself: existing techniques rely either on 
template-based generation or on the direct construction of a 
ranking relation from a particular trace. 
In contrast, our approach 
uses Craig interpolation to guide transition invariant synthesis 
with respect to the actual termination conditions, producing 
property-directed invariants.
%  from terminating traces. 



% There also exists another set of techniques that are based on the construction of well-founded transition
% invariants~\cite{DBLP:conf/lics/PodelskiR04,DBLP:conf/sas/CookPR05,DBLP:conf/cav/KroeningSTW10,DBLP:conf/tacas/TsitovichSWK11}.
% These approaches overapproximate the behaviour of the transition system, simplifying the termination proof.
% The core problem of these approaches is the generation of transition invariant.
% Listed techniques rely either on the template-based generation or on the direct generation of ranking relation based on a particular trace
% in the transition system.
% In contrast, we propose to do the guided generation of transition invariants, based on the terminating traces and the actual 
% termination conditions.
% Additionally, our technique is combined with a nontermination approach, increasing the efficiency of both techniques.




