\section{Conclusion}
\label{Sec:Conclusion}

% This paper introduces Interpolation-based Transition Invariant Generation for Proving and Disproving Termination.
% A novel approach for the termination analysis that extends Safety-based Nontermination Analysis technique.
% Our key insight is to use interpolation for the construction of termination witnesses,
% that allows the transition invariant to be guided by termination condition.
% Additionally, our technique is strongly centered on the interaction between termination and nontermination
% approaches, using transition invariants to limit the explored state space and using safety-based 
% nontermination to generate candidate transition invariants.
% Application of safety verifiers provides a strong background for the analysis, allowing to 
% quickly check termination and nontermination.
% We proved the soundness of the algorihtm and conducted an extensive experimental evaluation, 
% which demonstrates both the eficiency of the combination of termination and nontermination techniques,
% and the competitive  performance compared to other state-of-the-art termination analysis tools.

This paper presents an interpolation-based approach to transition 
invariant generation for proving and disproving termination of 
infinite-state systems. 
The approach extends safety-based nontermination analysis with a termination proving capability, 
unifying the two within a single framework. 
Our key insight is to exploit terminating traces produced by the SNA
as a source of structural information, using Craig interpolation 
to construct well-founded transition invariants guided by the 
termination conditions. 
The interaction between termination and nontermination reasoning is central to our technique: 
transition invariants are used to focus the nontermination search 
on non-covered states, while the SNA generates traces for invariants generation. 
Grounding the approach in safety verification provides a robust foundation, enabling 
efficient checking of both termination and nontermination via 
existing safety solvers. 
We proved the soundness of the algorithm and conducted an 
experimental evaluation on benchmarks from the Termination 
Competition. 
The results demonstrate both the benefit of combining 
termination and nontermination analysis and competitive performance 
against state-of-the-art termination analysis tools, including the 
unique solution of previously unsolved benchmark instances.



\textbf{Acknowledgements.} This work was supported by SNSF grant No.~XXXXXX.
We thank Marek Jankola and Dirk Beyer for fruitful discussions 
on termination analysis.